Reviewer Comments (Scientific Reports)

### 2020-03-23 notebook/sci_rep_reviews.Rmd
Revise response to reviewers, with line numbers, up to Reviewer 3

I have taken the following actions:

1. Re-run simulations while calculating the correlation between off-diagonal elements of A[i,j] and A[j,i] to see if the correlation changes with $\gamma$ across $S$. It does not. I also generated systems of known correlation coefficients to assess how this modulated $\gamma$'s affect on stability. It did not.

2. Run simulations across multiple structured network types to confirm the robustness of results. Network types include (a) cascade food webs, (b) small world networks, and (c) scale-free network.

3. Rewrite the manuscript, or the relevant parts of it.

@@

I again thank the reviewer for their comments. I agree that it is especially relevant that my manuscript focuses on finite S values, and I have made more emphasis on this point in my revision (line XXX).

@@

I agree with the relevance of Dougoud et al. (2018), and have made appropriate links to their results on feasibility. Nevertheless, one major result of my manuscript is that varying component response rates (i.e., $\gamma$) only affect stability, not feasibility. Hence, Reviewer 1 is incorrect that my conclusions concerning varying component response rates should be expected to enhance feasibility in experimentally observed food webs (and therefore that stability should be enhanced for biologically realistic systems in which all population densities are positive). Checking for validity in empirical food webs is therefore not entirely appropriate because the consequences of such an investigation would have negligible practical application.

I welcome Reviewer 1's call for expanding the scope of models considered for simulations, which was also recommended by Reviewer 2. Nevertheless, my manuscript focuses broadly on complex systems rather than pure ecological systems, so I have revised my manuscript to test more general models of common network structures. In terms of ecological models, in addition to random competition, mutualism, and predator-prey networks, I have now also included a cascade food web model (which is a foundation for the extended niche and nested-heirarchy models that Reviewer 1 mentions; Williams and Martinez 2000). Other network types that I have incorporated into my revision include small-world networks (Watts and Strogatz 1998) and scale-free networks (Albert and Barab\'{a}si 2002). Overall, the structured networks that I now include in the main text and supporting information encompass a wide range of physical, biological, and social systems.

More emphasis on feasibility and how it relates to ecological systems has now been included (line XXX). I have also added new sections in the Results (line XXX) and methods (line XXX) for the new simulations of structured networks. Full details of these results have also been added to the supporting information, and code for building structured networks is publicly available as a documented R package on GitHub (https://github.com/bradduthie/RandomMatrixStability).

@@

I have explained the result shown by the black line of Figures 3 (now 4), and the former figure 4 (now in the SI), now more clearly in my revision (line XXX).

@@

I appreciate these comments, and have tried to provide a more analytical explanation of the changes introduced by \gamma (line XXX). In making these changes, I have slightly adjusted the notation so that the text is more readable.

As Reviewer 2 correctly points out, my focus is on finite values of S and \sigma, in which complexity is gradually increased until it is dangerously close to the stability boundary. Given this focus, analytically demonstrating my central result (that variation in $X$ when $M = XA$ increases the stability of $M$), would require some sort of comparison between the probability distributions of eigenvalues for $M$ and $A$. For circular matrices in which elements are i.i.d ($A$), it has been proved that eigenvalues are distributed uniformly within a known boundary of support of the spectral density (Tao and Vu 2010). While recent papers have found the support for more general types of random matrices that include the form $M = AX$ (e.g, Gibbs et al. 2018; Neri and Metz 2012; Ahmadian et al. 2015), to the best of my knowledge, finding the actual density of eigenvalues within the support has not been attempted. This likely matters because for finite S, even though the real boundary of the support might be positive for both $M$ and $A$ (implying instability as $S \to \infty$; Grilli et al. 2017), the right most eigenvalue for a system still might be less than zero.

Reviewer 2 is correct that this is uncharted territory, and after carefully considering the possibility of analytical demonstration of my manuscript's results, I am convinced that such a demonstration is beyond the scope of this manuscript. My approach follows others with similar objectives. For example, Sinha (2005) also used a numerical approach to understand the stability of small-world networks (in contrast to the classic random networks of May).

@@

These were very helpful comments, and I am especially grateful to Reviewer 2 for doing a bit of preliminary exploration themselves to suggest improvements. I have now run simulations for different values of $\rho$, as suggested. Reviewer 2 is correct that the nonzero $\rho$ values do not change my results, but an analysis of stability across $\rho$ values also revealed an interesting pattern. While the relationship between $\rho$ and the leading eigenvalue is linear for circular matrices ($A$), it is convex when $\gamma$ is included such that $M = \gamma A$. I have now included these results into the main text of the manuscript, as I believe that they are important.

As suggested, I also considered matrices with even more structure. Specifically, I have simulated three broad classes of structured networks that are widespread in physical, social, and biological systems. These include food webs with cascade structures (e.g., Williams and Martinez 2000), small-world networks (Watts and Strogatz 1998), and scale-free networks (Albert and Barab\'{a}si 2002).

@@

I have revised the writing to more carefully frame the manuscript's key results (line XXX). This includes a change of the title, which now makes it more clear that the effects on stability are small while retaining an emphasis on generality across systems. I have also removed the word "markedly" from the abstract (line XXX).

>>>>

@@

I again thank the reviewer for their comments. I agree that it is especially relevant that my manuscript focuses on finite S values, and I have made more emphasis on this point in my revision (L48, L88, L209, L234). In particular, in addition to the statement on L48, I now state more explicitly in describing the methods that "the objective is to determine the effect of $\gamma$ as system complexity increases toward the boundary at which stability is realistic for a finite system." (L88)

@@

I agree with the relevance of Dougoud et al. (2018), and have made appropriate links to their results on feasibility (L143, L274). Nevertheless, one major result of my manuscript is that varying component response rates (i.e., $\gamma$) only affect stability, not feasibility (L148). Hence, Reviewer 1 is incorrect that my conclusions concerning varying component response rates should be expected to enhance feasibility in experimentally observed food webs (and therefore that stability should be enhanced for biologically realistic ecological communities in which all population densities are positive). Checking for validity in empirical food webs is therefore not entirely appropriate because the consequences of such an investigation would have negligible practical application.

I welcome Reviewer 1's call for expanding the scope of models considered for simulations, which was also recommended by Reviewer 2. Nevertheless, my manuscript focuses broadly on complex systems rather than only on pure ecological systems, so I have revised my manuscript to test more general models of common network structures. In terms of ecological models, in addition to random competition, mutualism, and predator-prey networks, I have now also included a cascade food web model (which is a foundation for the extended niche and nested-heirarchy models that Reviewer 1 mentions; Williams and Martinez 2000). Other network types that I have incorporated into my revision include small-world networks (Watts and Strogatz 1998) and scale-free networks (Albert and Barab\'{a}si 2002). Overall, the structured networks that I now include in the main text and supporting information encompass a wide range of physical, biological, and social systems. Results for thse simulations are presented in a new section entitled, "Simulations of structured M across S" (L123). Methods for constructing small-world (L247), scale-free (L260), and cascade food web (L268) networks are now explained, with additional details and figures now in the Supplemental Information. The results from these structured networks match those of the random networks; variation in $\gamma$ increases the number of stable systems given sufficiently high system complexity.

Feasibility and how it relates to ecological systems is explained on L141, with discussion starting on L188 for how this relates to ecological systems. Full details of these results have also been added to the supporting information, and code for building structured networks is publicly available as a documented R package on GitHub (https://github.com/bradduthie/RandomMatrixStability).

@@

I have explained the result shown by the black line of Figures 3 (now 4), and the former figure 4 (now in the SI), now more clearly in my revision (L417).

@@

I appreciate these comments, and have tried to provide a more analytical explanation of the changes introduced by \gamma (L53-L64), including a new subsection explaining the effect of variation in $\gamma$ on the covariation of M (L65-L81). I agree that these analytical results improve the manuscript, and I think that they are particularly useful in setting the stage for the numerical simulations that now follow. In making these changes, I have slightly adjusted the notation so that the text is more readable.

As Reviewer 2 correctly points out, my focus is on finite values of S and $\sigma$, in which complexity is gradually increased until it is dangerously close to the stability boundary. Given this focus, analytically demonstrating my central result (that variation in $X$ when $M = XA$ increases the probability of finding a stable $M$) would require some sort of comparison between the probability distributions of eigenvalues for $M$ and $A$. For circular matrices in which elements are i.i.d ($A$), it has been proved that eigenvalues are distributed uniformly within a known boundary of support of the spectral density (Tao and Vu 2010). While recent papers have found the support for more general types of random matrices that include the form $M = AX$ (e.g, Gibbs et al. 2018; Neri and Metz 2012; Ahmadian et al. 2015), to the best of my knowledge, finding the actual density of eigenvalues within the support has not been attempted. This likely matters because for finite S, even though the real boundary of the support might be positive for both $M$ and $A$ (implying instability as $S \to \infty$; Grilli et al. 2017), the right-most eigenvalue for a system still might be less than zero.

Reviewer 2 is correct that this is uncharted territory, and after spending a lot of time considering the possibility of an analytical demonstration of my main results, I am convinced that such a demonstration is beyond the scope of this manuscript. My approach follows others with similar objectives. For example, Sinha (2005) also used a numerical approach to understand the stability of small-world networks (in contrast to the classic random networks of May).

@@

These were very helpful comments, and I have embraced them wholeheartedly in my revision. I am especially grateful to Reviewer 2 for doing a bit of preliminary exploration themselves to suggest improvements. I have now included a full section devoted to the role of $\rho$ on stability, and run simulations for different values of $\rho$, as suggested. Reviewer 2 is correct that the nonzero $\rho$ values do not change my results, but an analysis of stability across $\rho$ values also revealed an interesting pattern. While the relationship between $\rho$ and the leading eigenvalue is linear for circular matrices ($A$), it is convex when $\gamma$ is included such that $M = \gamma A$. I have now included these results into the main text of the manuscript, as I believe that they are important.

As suggested, I also considered matrices with even more structure, performing a new simulations on each. Specifically, I have simulated three broad classes of structured networks that are widespread in physical, social, and biological systems. These include food webs with cascade structures (e.g., Williams and Martinez 2000), small-world networks (Watts and Strogatz 1998), and scale-free networks (Albert and Barab\'{a}si 2002). Results for these networks are presented in L123-L140, and methods are explained in L247-L273, and in the Supplemental Information. Overall, these structured networks were affected by variation in $\gamma$ in the same way as random networks (increasing stability given sufficiently high complexity).

@@

As recommended, I have changed the title, which now makes it more clear that the effects on stability are restricted to highly complex finite systems, while still retaining an emphasis on generality across systems. I have also removed the word "markedly" from the abstract, as recommended.
